Ubiquitous Digital Technologies and Spatial Structure; an update

### 2020-12-03 paper/2_resubmission/tranos_ioannides_unmarked.tex
unmarked

\color{blue}

Kolko {[}22{]} uses internet diffusion data and identifies a clear
complementary link between internet usage and city size. Interestingly,
he identifies higher internet domain densities in remote cities which
indicates a substitution effect of the internet for longer-distance
non-electronic communications. His results are consistent for different
measures of internet diffusion (internet domain density and internet
take-up). Sinai and Waldfogel {[}23{]} approach the same question from
the consumers' point of view and find support for complementarity
between internet and cities. They study the link between market size and
locally targeted online content and find more online local content in
larger markets. However, they also obtain evidence for a substitution
effect: holding local online content constant, market size has a
negative effect on individual connectivity. Forman et al.~{[}24{]}
examines whether commercial internet adoption is higher in cities than
in rural areas. While the former would indicate a complementarity
between internet adoption and cities, the latter would reflect a
substitutability relation, according to which the internet is used as a
means to offset costs and lack of opportunities related to
peripheriality. Their results indicate that despite internet adoption by
firms with more than 100 employees being greater in smaller urban
agglomerations, the adoption of more sophisticated internet-based
applications is positively related with city size in \(2000\). Sohn,
Kim, and Hewings {[}25{]} compare how information technologies are
related to urban spatial structure for Chicago and Seoul. Although they
find a clear complementary link for Chicago, this is not the case for
Seoul, where information technologies contribute to a more dispersed
spatial pattern. Focusing on the municipalities in the province of
Barcelona, Pons-Novell and Viladecans-Marsal {[}26{]} find a
complementary link between individual internet take-up and off-line
commercial offerings, but their results cannot safely reject a
substitution effect. Craig, Hoang, and Kohlhase {[}27{]} focus on
internet take-up rates across US states during the period \(2000-2011\).
Their analysis provides suggestive evidence of a complementary role that
internet connectivity performs on urban living. Ioannides et
al.~{[}18{]} use country-level data on city size distributions to
examine the impact of fixed line telephony on urban structure
\(1980-2000\). Using a panel dataset of spatial dispersion measures,
they find robust evidence that an increase in the number of telephone
lines per capita encourages the spatial dispersion of population in that
they lead to a more concentrated distribution of city sizes. In
addition, by the end of their study period the internet has come into
use, but their results with internet usage is more speculative but do
show that it goes in the same direction.

\color{blue}

Focusing on rural areas, Partridge et al.~{[}28{]} find no evidence that
rural distance penalties in the US have substantially changed since
\(1970s\) indicating that technological changes including the internet
and digital communications have not affected spatial structure. Also
interesting are the findings of a more recent study by Kim and Orazem
{[}29{]} on the economic effects of broadband internet in rural areas.
They identify a positive effect on new firm location decisions, with the
effect being higher in rural areas with larger population and in rural
areas adjacent to a metropolitan area. These results suggest a
complementary relation between the internet and agglomeration economies.

\color{blue}

Moving beyond modelling the direct relationship between ICT and spatial
structure, Bekkerman and Gilpin {[}30{]} focused on the role of locally
based information resources using a dataset about the US libraries
during the period \(2000-2008\). Their results suggest that internet
access increases the demand and the value of locally accessible
information, and such complementarities are higher in larger
metropolitan areas. Anenberg and Kung {[}31{]} also identified a
complementary relation between the internet and consumption variety in
cities by focusing on food truck industry in the US.

@@

\color{blue}

@@

The multi-country identification strategy is a two-step approach
{[}18{]}. The first step of our methodology is to estimate the
\color{blue} Pareto exponent \color{black} for a broad sample of
countries over time. \color{blue} The Pareto exponent \color{black} is
one of the most widely used measures of spatial dispersion with numerous
applications in urban economics and economic geography ({[}see for
example {[}37{]}; {[}38{]}; {[}39{]}; {[}40{]}; {[}41{]}; {[}42{]};
{[}43{]}). This an appropriate measure of dispersion because of the
extreme heterogeneity of the city size distribution and the very good
fits normally obtained with such estimations\footnote{\color{blue} We
  also replicate our analysis using other measures of dispersion in
  Appendix S1. Our results remain consistent.}. \color{black} Skipping
details pertaining of the suitability of the approach, which may be
found in Ioannides et al.~{[}18{]} and Tranos and Ioannides {[}19{]}, we
proceed with the estimation of the logarithmic form of:

@@

\noindent The model enable us to estimate the effects of a vector \(X\)
of explanatory variables which are included to account for the spatial
structure of country \(c\) in year \(t\). The main variables of interest
here are internet and digital communications variables that include:
internet users per \(100\) inhabitants, broadband users per \(100\)
inhabitants, mobile phone users per \(100\) inhabitants, and fixed phone
users per \(100\) inhabitants. To address a potential omitted variable
bias equation \ref{eq3} includes \color{blue} country (\(\theta_c\)) and
year (\(\delta_t\)) fixed effects; \color{black} \(e_{ct}\) is the error
term. In addition, vector \(X\) includes several control variables,
whose descriptive statistics together with those for the other variables
used to estimate \ref{eq3} are reported in Table \ref{desc.global}.

@@

This section presents the estimation results of \ref{eq3}. The LHS
variable is \(\zeta\), the \color{blue} Pareto exponent, \color{black}
as estimated according to the Gabaix and Ibragimov {[}45{]} correction.
The main explanatory variables of interest are provided at the top of
Table \ref{ols.global}; namely internet, broadband, mobile and fixed
telephony per \(100\) habitants, expressed in natural logarithms. The
main variables of interest are introduced successively on their own in
the regressions reported in Table \ref{ols.global}. All regressions
include country fixed effects to control for unobserved heterogeneity
and a time trend. In addition, the observations are weighted with the
inverse squared standard error of the estimated \color{blue} Pareto
exponent \color{black} (see Table \ref{desc.global}) to address
potential noise that is carried over from the first part of our
identification strategy. In regards to the interpretation of the
estimated coefficients, given that the \color{blue} Pareto exponent
\color{black} has entered the regression not as an absolute value, but
instead as a real number a positive coefficient for a RHS variable
indicates an impact towards the decrease of the spatial dispersion of
population. In other words, a positive coefficient indicates an effect
towards less uniform city sizes that is more dispersion of city sizes.
The latter is indicative of enhancement of agglomeration economies
because of the expansion of digital technologies.

@@

Table \ref{ols.global} and \ref{2sls.global} report the estimation
results based on the UN urban agglomeration data. Starting with the
former, we note that an agglomerative effect is only detected for the
internet users and while a marginally opposing result can be seen for
broadband users as indicated by the significant coefficients in columns
(1) and (2). For the telephony variables the estimation of \ref{eq3} did
not yield statistically significant coefficients. Before discussing
further these results and the effects of the other control variables, we
need to highlight that the main challenge of estimating equation
\ref{eq3} is the potential endogenous nature of the share of internet
users which might prevent us from being able to infer a truly causal
effect. Endogeneity might be an issue here as spatial structure, which
is proxied by the \color{blue} Pareto exponent, \color{black} might be
affected by another source, which also affects internet penetration. For
instance, economic development might affect the concentration of
population in large cities and at the same time enable more people to go
online. If we do not address this issue, the coefficient for the main
variable of interest will capture potential effects that internet
penetration has on spatial structure, but also potential reverse
causality effects that spatial structure might generate on internet
penetration. To overcome this potential problem, Table \ref{2sls.global}
reports estimates of equation \ref{eq3} using two-stage least squares
(2SLS) with instrumental variables (IVs). The latter are variables which
are correlated with our endogenous variables, but do not influence
current spatial structure. Such an approach will enable us to estimate
the causal effect -- if any -- of the internet and digital
communications on spatial structure. At a first stage, our endogenous
variables are regressed against the IVs. Then, the predicted values of
the endogenous variables based on the IVs and the other control
variables are used instead of the endogenous variable to estimate eq.
\ref{eq3}. A significant effect will verify the causal impact of the
internet and digital communication usage on spatial structure.

@@

The main challenge for such an exercise is to find a valid IV. In our
case, the challenge is even bigger as we need to find a variable which
is not only fit for purpose, but it also contain as few missing values
as possible in order to retain the same number of observations as the
OLS models. This is particularly difficult for a diverse sample of
countries as the one we are dealing with here. We propose here the
percentage of female participation in the labor force, which is directly
related to digital infrastructure, but not to spatial structure (see
Table \ref{desc.global} for the descriptive statistics). \color{blue}
The literature provides empirical and theoretical evidence that ICT are
correlated with female participation in labour markets mostly because
they enable teleworking {[}58--62{]} \color{black} At the same time,
female labor participation is not related with spatial structure
especially given the diverse sample of countries. Although one might
think that increased female labor participation may lead to relocation
of households to large cities and therefore affect the spatial structure
of cities, such a process varies a lot within our dataset. Even if this
might be true for developed countries with mature urban systems for
which a location within a large urban center might provide opportunities
for both male and female workers to find jobs, our dataset also includes
countries, \(41\) per cent of the GDP of which is attributed to primary
activities (see Table \ref{desc.global}). So, female labor participation
might also be related to economic activities located outside large urban
centers. In terms of the relevant tests presented in Table
\ref{2sls.global} the weak identification first stage F-test exceeds by
far all the Stock-Yogo critical values for all, but the mobile telephony
regression. Hence, we are confident to interpret the effects of
internet, broadband internet and fixed telephony usage.

The estimations presented in Table \ref{2sls.global} indicate that
increases in internet and broadband internet usage have resulted to a
decrease of the spatial dispersion of population for the time period and
for the panel of countries included in our data. In other words,
increases in internet and digital communications resulted to national
urban systems which are less uniform in terms of city sizes and are
characterized by higher population concentrations in larger cities. The
exact opposite appears to happen for fixed telephony, which our results
indicate that has led to an increase of spatial dispersion of
population. \color{blue} We expect that the IV estimates capture the
local average treatment effect {[}63{]} of these labour markets, which
are characterised by more active engagement with the internet, and,
therefore, experience a larger effect on the Pareto exponents.
\color{black} Importantly, our identification strategy enables us to
address potential reverse causality issues and treat the results of
Table \ref{2sls.global} as causal. Hence, the main finding of our
multi-country, global analysis based on data for urban agglomerations of
at least 300,000 inhabitants is that internet and broadband internet
usage appear to act in favor of agglomeration economies and result to
urban systems with more dominant cities on the top of the urban
hierarchy. On the contrary, fixed telephony usage acted in favour of
centrifugal forces. Our identification strategy does not enable us to
make similar interpretations for mobile telephony usage.

@@

In total, the results indicate that our measures of internet and
telephony penetration have further enhanced agglomeration forces, at
least for large urban agglomerations, while the opposite happens for
fixed telephony. These results are robust against endogeneity issues,
but are limited to urban agglomeration included in our data. Indeed, for
the estimation of the \color{blue} Pareto exponent \color{black} we only
included urban agglomeration of \(300,000\) inhabitants or more due to
data availability. Hence, the above estimations cannot verify whether
such an effect is also valid for smaller cities. Interestingly, the
above results regarding the role of fixed telephony are in agreement
with previous results from {[}18{]} {[}19{]} which used a smaller panel
of countries but without constraints regarding the city size. However,
our results do not support the findings of {[}19{]} regarding internet
usage.

\color{blue}

@@

The LHS variable we use here is the outcome of equation \ref{eq5}.
\color{blue} We use this variable to estimate \ref{eq4} with OLS.
Additionally, we employ a normalized form of the dependent variable, so
as that variable is bounded between \(0\) and \(1\) and with a mean of
\(0.5\)\footnote{\[Normalized\:difference\:in\:ranks_{i} = \frac{difference\:in\:ranks_{i} + N_{cities}}{2xN_{cities}}.\]}.
This transformation enables us to estimate \ref{eq4} with a
quasibinomial GLM estimator given that the original form of our LHS
variable does not vary continuously, but is instead defined as a
difference between two count variables, which may also assume negative
values. \color{blue} The results of the different estimations remain
qualitatively the same regardless of the form of the LHS variable we
use. Table \ref{ols.usa} reports the results of the estimation of
equation \ref{eq4} using OLS. The sign and significance of the main
variable of interest -- percentage of population with computer and
broadband connection -- verifies our global level results. That is, an
increase in internet usage improves the position of a city in the US
urban system. The results are consistent among the different estimators
that is OLS and GLM. In addition, Table \ref{ols.usa} also reports the
estimates of interaction effects between the share of population with
broadband connection with population and population density. Although
the interaction terms are not significant, the main effect remains
qualitatively unchanged.

@@

Although not directly comparable, the above estimations are in
accordance with our global model. Moreover, the potential presence of
endogeneity might be a problem here as well. Therefore, Table
\ref{2sls.usa} reports 2SLS estimations. In addition to the strategy
employed for the cross-country analysis -- that is the inclusion of one
instrument for which we have strong reasons to believe is uncorrelated
with the error term -- data availability allows us to employ a second IV
in order to estimate the Sargan over-identification restrictions test
(Column 2). The main instrument we propose here is Bachelor's degrees
per inhabitant in \(2005\)\footnote{\color{blue} Due to differences in
  scale, time and the endogenous variables, different IVs have been
  unutilized for the different sections of the paper. \color{black}}.
Even if the quality of human capital affect the population growth of a
city \(10\) years later, our LHS variable adopts a systemic
understanding of the US urban system as it measures the relative
position of a city within the overall urban system instead of its
population growth. In other words, a city might experience population
growth between two periods, but if other cities have also experienced
population growth, this might not affect its relative position within
the urban system. Hence, we do not expect that this IV affects the LHS
variable. Moreover, when we add the second IV, which in this case is the
commute time in minutes also in \(2005\), we fail to reject the null
hypothesis of the Sargan test, something which also adds the validity of
our strategy. Furthermore, our results do not suffer from weak
identification according to the relevant tests in Table \ref{2sls.usa}.

@@

In total, the estimates of Table \ref{2sls.usa} enable us to identify a
causal effect of the share of households with internet connection on the
position of a city in the urban system. Specifically, an increase in
internet penetration led to an improvement of the position of a city in
the US urban hierarchy for the cities and the time period included in
our analysis. \color{blue} Everything else being equal, if a city had
experienced an increase of 10 per cent of internet penetration, this
increase would have improved its relative position by \(9\) places.
\color{black} As in the global model reported in
\protect\hyperlink{sec4}{Section 4}, internet penetration appears to
work in favor of agglomeration externalities. In regards to the control
variables, we can identify a negative effect of income, which is
consistent with the effect of GDP per capita in the global model
presented in Table \ref{2sls.global}. During the study period,
unemployment rates and the share of white population negatively affected
the ranking of micropolitan and metropolitan areas in the US as
reflected in the relevant regressions.

@@

In regards to IV strategy, we use here the \emph{number of universities}
as the main instrument. Universities in England and Wales have been
established long time ago and therefore we do not expect them to affect
the change in the relative position of built-up areas in \(2011\).
\color{blue} The presence of universities and large student populations
tend to be correlated with the quality of internet infrastructure
provision {[}71{]}. \color{black} The addition of the second instrument
(absolute number of broadband tests in 2011) in column \(2\) results to
a Sargan test, the null hypothesis of which cannot be rejected.
Moreover, the estimations reported in Table 11 do not suffer from weak
instruments as the relevant test is above the rule of thumb value of
\(10\). \color{blue} The coefficients derived from OLS (Table
\ref{ols.uk}) and 2SLS (Table \ref{2sls.uk}) are always positive and
significant, something which advocates in favour of the effect of
broadband speed in enhancing agglomeration forces. Interestingly, the
magnitude of the 2SLS coefficient (column 1 in Table \ref{2sls.uk}) is
almost 8 times higher than the OLS coefficient (column 1 in Table
\ref{ols.uk}). We attribute the OLS underestimation, which is more or
less consistent in all our estimations, to the difficulty in directly
observing active engagement with ICT. Our endogenous variable captures
the quality of the internet infrastructure as the experienced internet
speed. Similarly to the endogenous variables of the global and the US
model (internet and telephony take-up), what these variables do not
capture is what individuals actually do with ICT {[}72{]}. For example,
if we assume a bi-modal internet usage for either entertainment or work
purposes, one would expect that the latter would have a larger effect on
spatial structure than the former. We believe that our endogenous
variables capture mostly the former. Therefore, the IV estimations,
especially for UK case, reveal the local average treatment effect
{[}63{]} by capturing the effect of these users/households who actively
engage with ICT (e.g.~work on ICT-related sectors) and, therefore, have
a higher effect on spatial structure. A back of the envelope calculation
indicates that, everything else being equal, \(10\) per cent increase in
download speed for a BUA, would have increased its position by \(4\)
places. \color{black}

>>>>

--\textgreater{} Kolko {[}22{]} uses internet diffusion data and
identifies a clear complementary link between internet usage and city
size. Interestingly, he identifies higher internet domain densities in
remote cities which indicates a substitution effect of the internet for
longer-distance non-electronic communications. His results are
consistent for different measures of internet diffusion (internet domain
density and internet take-up). Sinai and Waldfogel {[}23{]} approach the
same question from the consumers' point of view and find support for
complementarity between internet and cities. They study the link between
market size and locally targeted online content and find more online
local content in larger markets. However, they also obtain evidence for
a substitution effect: holding local online content constant, market
size has a negative effect on individual connectivity. Forman et
al.~{[}24{]} examines whether commercial internet adoption is higher in
cities than in rural areas. While the former would indicate a
complementarity between internet adoption and cities, the latter would
reflect a substitutability relation, according to which the internet is
used as a means to offset costs and lack of opportunities related to
peripheriality. Their results indicate that despite internet adoption by
firms with more than 100 employees being greater in smaller urban
agglomerations, the adoption of more sophisticated internet-based
applications is positively related with city size in \(2000\). Sohn,
Kim, and Hewings {[}25{]} compare how information technologies are
related to urban spatial structure for Chicago and Seoul. Although they
find a clear complementary link for Chicago, this is not the case for
Seoul, where information technologies contribute to a more dispersed
spatial pattern. Focusing on the municipalities in the province of
Barcelona, Pons-Novell and Viladecans-Marsal {[}26{]} find a
complementary link between individual internet take-up and off-line
commercial offerings, but their results cannot safely reject a
substitution effect. Craig, Hoang, and Kohlhase {[}27{]} focus on
internet take-up rates across US states during the period \(2000-2011\).
Their analysis provides suggestive evidence of a complementary role that
internet connectivity performs on urban living. Ioannides et
al.~{[}18{]} use country-level data on city size distributions to
examine the impact of fixed line telephony on urban structure
\(1980-2000\). Using a panel dataset of spatial dispersion measures,
they find robust evidence that an increase in the number of telephone
lines per capita encourages the spatial dispersion of population in that
they lead to a more concentrated distribution of city sizes. In
addition, by the end of their study period the internet has come into
use, but their results with internet usage is more speculative but do
show that it goes in the same direction.

--\textgreater{} Focusing on rural areas, Partridge et al.~{[}28{]} find
no evidence that rural distance penalties in the US have substantially
changed since \(1970s\) indicating that technological changes including
the internet and digital communications have not affected spatial
structure. Also interesting are the findings of a more recent study by
Kim and Orazem {[}29{]} on the economic effects of broadband internet in
rural areas. They identify a positive effect on new firm location
decisions, with the effect being higher in rural areas with larger
population and in rural areas adjacent to a metropolitan area. These
results suggest a complementary relation between the internet and
agglomeration economies.

--\textgreater{} Moving beyond modelling the direct relationship between
ICT and spatial structure, Bekkerman and Gilpin {[}30{]} focused on the
role of locally based information resources using a dataset about the US
libraries during the period \(2000-2008\). Their results suggest that
internet access increases the demand and the value of locally accessible
information, and such complementarities are higher in larger
metropolitan areas. Anenberg and Kung {[}31{]} also identified a
complementary relation between the internet and consumption variety in
cities by focusing on food truck industry in the US.

@@

@@

The multi-country identification strategy is a two-step approach
{[}18{]}. The first step of our methodology is to estimate the Pareto
exponent \color{black} for a broad sample of countries over time. The
Pareto exponent \color{black} is one of the most widely used measures of
spatial dispersion with numerous applications in urban economics and
economic geography ({[}see for example {[}37{]}; {[}38{]}; {[}39{]};
{[}40{]}; {[}41{]}; {[}42{]}; {[}43{]}). This an appropriate measure of
dispersion because of the extreme heterogeneity of the city size
distribution and the very good fits normally obtained with such
estimations\footnote{We also replicate our analysis using other measures
  of dispersion in Appendix S1. Our results remain consistent.}.
\color{black} Skipping details pertaining of the suitability of the
approach, which may be found in Ioannides et al.~{[}18{]} and Tranos and
Ioannides {[}19{]}, we proceed with the estimation of the logarithmic
form of:

@@

\noindent The model enable us to estimate the effects of a vector \(X\)
of explanatory variables which are included to account for the spatial
structure of country \(c\) in year \(t\). The main variables of interest
here are internet and digital communications variables that include:
internet users per \(100\) inhabitants, broadband users per \(100\)
inhabitants, mobile phone users per \(100\) inhabitants, and fixed phone
users per \(100\) inhabitants. To address a potential omitted variable
bias equation \ref{eq3} includes country (\(\theta_c\)) and year
(\(\delta_t\)) fixed effects; \color{black} \(e_{ct}\) is the error
term. In addition, vector \(X\) includes several control variables,
whose descriptive statistics together with those for the other variables
used to estimate \ref{eq3} are reported in Table \ref{desc.global}.

@@

This section presents the estimation results of \ref{eq3}. The LHS
variable is \(\zeta\), the Pareto exponent, \color{black} as estimated
according to the Gabaix and Ibragimov {[}45{]} correction. The main
explanatory variables of interest are provided at the top of Table
\ref{ols.global}; namely internet, broadband, mobile and fixed telephony
per \(100\) habitants, expressed in natural logarithms. The main
variables of interest are introduced successively on their own in the
regressions reported in Table \ref{ols.global}. All regressions include
country fixed effects to control for unobserved heterogeneity and a time
trend. In addition, the observations are weighted with the inverse
squared standard error of the estimated Pareto exponent \color{black}
(see Table \ref{desc.global}) to address potential noise that is carried
over from the first part of our identification strategy. In regards to
the interpretation of the estimated coefficients, given that the Pareto
exponent \color{black} has entered the regression not as an absolute
value, but instead as a real number a positive coefficient for a RHS
variable indicates an impact towards the decrease of the spatial
dispersion of population. In other words, a positive coefficient
indicates an effect towards less uniform city sizes that is more
dispersion of city sizes. The latter is indicative of enhancement of
agglomeration economies because of the expansion of digital
technologies.

@@

Table \ref{ols.global} and \ref{2sls.global} report the estimation
results based on the UN urban agglomeration data. Starting with the
former, we note that an agglomerative effect is only detected for the
internet users and while a marginally opposing result can be seen for
broadband users as indicated by the significant coefficients in columns
(1) and (2). For the telephony variables the estimation of \ref{eq3} did
not yield statistically significant coefficients. Before discussing
further these results and the effects of the other control variables, we
need to highlight that the main challenge of estimating equation
\ref{eq3} is the potential endogenous nature of the share of internet
users which might prevent us from being able to infer a truly causal
effect. Endogeneity might be an issue here as spatial structure, which
is proxied by the Pareto exponent, \color{black} might be affected by
another source, which also affects internet penetration. For instance,
economic development might affect the concentration of population in
large cities and at the same time enable more people to go online. If we
do not address this issue, the coefficient for the main variable of
interest will capture potential effects that internet penetration has on
spatial structure, but also potential reverse causality effects that
spatial structure might generate on internet penetration. To overcome
this potential problem, Table \ref{2sls.global} reports estimates of
equation \ref{eq3} using two-stage least squares (2SLS) with
instrumental variables (IVs). The latter are variables which are
correlated with our endogenous variables, but do not influence current
spatial structure. Such an approach will enable us to estimate the
causal effect -- if any -- of the internet and digital communications on
spatial structure. At a first stage, our endogenous variables are
regressed against the IVs. Then, the predicted values of the endogenous
variables based on the IVs and the other control variables are used
instead of the endogenous variable to estimate eq. \ref{eq3}. A
significant effect will verify the causal impact of the internet and
digital communication usage on spatial structure.

@@

The main challenge for such an exercise is to find a valid IV. In our
case, the challenge is even bigger as we need to find a variable which
is not only fit for purpose, but it also contain as few missing values
as possible in order to retain the same number of observations as the
OLS models. This is particularly difficult for a diverse sample of
countries as the one we are dealing with here. We propose here the
percentage of female participation in the labor force, which is directly
related to digital infrastructure, but not to spatial structure (see
Table \ref{desc.global} for the descriptive statistics). The literature
provides empirical and theoretical evidence that ICT are correlated with
female participation in labour markets mostly because they enable
teleworking {[}58--62{]} \color{black} At the same time, female labor
participation is not related with spatial structure especially given the
diverse sample of countries. Although one might think that increased
female labor participation may lead to relocation of households to large
cities and therefore affect the spatial structure of cities, such a
process varies a lot within our dataset. Even if this might be true for
developed countries with mature urban systems for which a location
within a large urban center might provide opportunities for both male
and female workers to find jobs, our dataset also includes countries,
\(41\) per cent of the GDP of which is attributed to primary activities
(see Table \ref{desc.global}). So, female labor participation might also
be related to economic activities located outside large urban centers.
In terms of the relevant tests presented in Table \ref{2sls.global} the
weak identification first stage F-test exceeds by far all the Stock-Yogo
critical values for all, but the mobile telephony regression. Hence, we
are confident to interpret the effects of internet, broadband internet
and fixed telephony usage.

The estimations presented in Table \ref{2sls.global} indicate that
increases in internet and broadband internet usage have resulted to a
decrease of the spatial dispersion of population for the time period and
for the panel of countries included in our data. In other words,
increases in internet and digital communications resulted to national
urban systems which are less uniform in terms of city sizes and are
characterized by higher population concentrations in larger cities. The
exact opposite appears to happen for fixed telephony, which our results
indicate that has led to an increase of spatial dispersion of
population. We expect that the IV estimates capture the local average
treatment effect {[}63{]} of these labour markets, which are
characterised by more active engagement with the internet, and,
therefore, experience a larger effect on the Pareto exponents.
\color{black} Importantly, our identification strategy enables us to
address potential reverse causality issues and treat the results of
Table \ref{2sls.global} as causal. Hence, the main finding of our
multi-country, global analysis based on data for urban agglomerations of
at least 300,000 inhabitants is that internet and broadband internet
usage appear to act in favor of agglomeration economies and result to
urban systems with more dominant cities on the top of the urban
hierarchy. On the contrary, fixed telephony usage acted in favour of
centrifugal forces. Our identification strategy does not enable us to
make similar interpretations for mobile telephony usage.

@@

In total, the results indicate that our measures of internet and
telephony penetration have further enhanced agglomeration forces, at
least for large urban agglomerations, while the opposite happens for
fixed telephony. These results are robust against endogeneity issues,
but are limited to urban agglomeration included in our data. Indeed, for
the estimation of the Pareto exponent \color{black} we only included
urban agglomeration of \(300,000\) inhabitants or more due to data
availability. Hence, the above estimations cannot verify whether such an
effect is also valid for smaller cities. Interestingly, the above
results regarding the role of fixed telephony are in agreement with
previous results from {[}18{]} {[}19{]} which used a smaller panel of
countries but without constraints regarding the city size. However, our
results do not support the findings of {[}19{]} regarding internet
usage.

@@

The LHS variable we use here is the outcome of equation \ref{eq5}. We
use this variable to estimate \ref{eq4} with OLS. Additionally, we
employ a normalized form of the dependent variable, so as that variable
is bounded between \(0\) and \(1\) and with a mean of \(0.5\)\footnote{\[Normalized\:difference\:in\:ranks_{i} = \frac{difference\:in\:ranks_{i} + N_{cities}}{2xN_{cities}}.\]}.
This transformation enables us to estimate \ref{eq4} with a
quasibinomial GLM estimator given that the original form of our LHS
variable does not vary continuously, but is instead defined as a
difference between two count variables, which may also assume negative
values. The results of the different estimations remain qualitatively
the same regardless of the form of the LHS variable we use. Table
\ref{ols.usa} reports the results of the estimation of equation
\ref{eq4} using OLS. The sign and significance of the main variable of
interest -- percentage of population with computer and broadband
connection -- verifies our global level results. That is, an increase in
internet usage improves the position of a city in the US urban system.
The results are consistent among the different estimators that is OLS
and GLM. In addition, Table \ref{ols.usa} also reports the estimates of
interaction effects between the share of population with broadband
connection with population and population density. Although the
interaction terms are not significant, the main effect remains
qualitatively unchanged.

@@

Although not directly comparable, the above estimations are in
accordance with our global model. Moreover, the potential presence of
endogeneity might be a problem here as well. Therefore, Table
\ref{2sls.usa} reports 2SLS estimations. In addition to the strategy
employed for the cross-country analysis -- that is the inclusion of one
instrument for which we have strong reasons to believe is uncorrelated
with the error term -- data availability allows us to employ a second IV
in order to estimate the Sargan over-identification restrictions test
(Column 2). The main instrument we propose here is Bachelor's degrees
per inhabitant in \(2005\)\footnote{Due to differences in scale, time
  and the endogenous variables, different IVs have been unutilized for
  the different sections of the paper. \color{black}}. Even if the
quality of human capital affect the population growth of a city \(10\)
years later, our LHS variable adopts a systemic understanding of the US
urban system as it measures the relative position of a city within the
overall urban system instead of its population growth. In other words, a
city might experience population growth between two periods, but if
other cities have also experienced population growth, this might not
affect its relative position within the urban system. Hence, we do not
expect that this IV affects the LHS variable. Moreover, when we add the
second IV, which in this case is the commute time in minutes also in
\(2005\), we fail to reject the null hypothesis of the Sargan test,
something which also adds the validity of our strategy. Furthermore, our
results do not suffer from weak identification according to the relevant
tests in Table \ref{2sls.usa}.

@@

In total, the estimates of Table \ref{2sls.usa} enable us to identify a
causal effect of the share of households with internet connection on the
position of a city in the urban system. Specifically, an increase in
internet penetration led to an improvement of the position of a city in
the US urban hierarchy for the cities and the time period included in
our analysis. Everything else being equal, if a city had experienced an
increase of 10 per cent of internet penetration, this increase would
have improved its relative position by \(9\) places. \color{black} As in
the global model reported in \protect\hyperlink{sec4}{Section 4},
internet penetration appears to work in favor of agglomeration
externalities. In regards to the control variables, we can identify a
negative effect of income, which is consistent with the effect of GDP
per capita in the global model presented in Table \ref{2sls.global}.
During the study period, unemployment rates and the share of white
population negatively affected the ranking of micropolitan and
metropolitan areas in the US as reflected in the relevant regressions.

@@

In regards to IV strategy, we use here the \emph{number of universities}
as the main instrument. Universities in England and Wales have been
established long time ago and therefore we do not expect them to affect
the change in the relative position of built-up areas in \(2011\). The
presence of universities and large student populations tend to be
correlated with the quality of internet infrastructure provision
{[}71{]}. \color{black} The addition of the second instrument (absolute
number of broadband tests in 2011) in column \(2\) results to a Sargan
test, the null hypothesis of which cannot be rejected. Moreover, the
estimations reported in Table 11 do not suffer from weak instruments as
the relevant test is above the rule of thumb value of \(10\). The
coefficients derived from OLS (Table \ref{ols.uk}) and 2SLS (Table
\ref{2sls.uk}) are always positive and significant, something which
advocates in favour of the effect of broadband speed in enhancing
agglomeration forces. Interestingly, the magnitude of the 2SLS
coefficient (column 1 in Table \ref{2sls.uk}) is almost 8 times higher
than the OLS coefficient (column 1 in Table \ref{ols.uk}). We attribute
the OLS underestimation, which is more or less consistent in all our
estimations, to the difficulty in directly observing active engagement
with ICT. Our endogenous variable captures the quality of the internet
infrastructure as the experienced internet speed. Similarly to the
endogenous variables of the global and the US model (internet and
telephony take-up), what these variables do not capture is what
individuals actually do with ICT {[}72{]}. For example, if we assume a
bi-modal internet usage for either entertainment or work purposes, one
would expect that the latter would have a larger effect on spatial
structure than the former. We believe that our endogenous variables
capture mostly the former. Therefore, the IV estimations, especially for
UK case, reveal the local average treatment effect {[}63{]} by capturing
the effect of these users/households who actively engage with ICT
(e.g.~work on ICT-related sectors) and, therefore, have a higher effect
on spatial structure. A back of the envelope calculation indicates that,
everything else being equal, \(10\) per cent increase in download speed
for a BUA, would have increased its position by \(4\) places.
\color{black}
